The Morality of Arbitrage

### 2017-06-15 Slides/apee-slides.Rmd
Changes for the APEE presentation.

---
title: "The Morality of Arbitrage - Some Half-Baked Ideas"
author: "Tyler J. Brough"
date: "September 23, 2016"
output: ioslides_presentation
---

#

@@

@@

>- It is apparent to me that the standard ethical arguments for capitalism are not having any purchase in the minds of most people.
>- Capitalism is presented as some kind of Faustian bargain: we like it for its wealth production and efficiency, but at bottom it is really about greed. It is rapacious and immoral.
>- It is readily apparent that many of our colleagues in the business disciplines believe this notion
>- __I'm afraid that our students do too!__ They leave here with some good technical tools, but not with an appreciation of just how much good they will do in their careers. 
>- That is on us! That's a failure on our part to teach correct principles.

## Some Evidence

I asked my finance bros (and sis) for examples of the "evils" of finance and capitalism:

From Paul:

@@

From Ben:

>- "Derivatives are financial weapons of mass destruction" - Warren Buffett
>- "Derivatives are toxic - stay away from them" - unnamed donor to the Huntsman School
>- "In a world where profit is consistently put before both people and the planet, climate economics has everything to do with ethics and morality" - Naomi Klein

## Some Evidence Continued

From Ryan:

>- "I heard a quote on the radio today that 46% of millenials don't think capitalism works."
>- "The Occupy Wall Street protests are shining a national spotlight on the most powerful, dangerous and secretive economic and political force in America." --Bernie Sanders
>- "What Wall Street and credit card companies are doing is really not much different from what gangsters and loan sharks do who make predatory loans. While the bankers wear three-piece suits and don't break the knee caps of those who can't pay back, they still are destroying people's lives." --Bernie Sanders

## Yet More Evidence

From Dani:

>-  JF from 2015 by Zingales is [*Does Finance Benefit Society?*](http://faculty.chicagobooth.edu/luigi.zingales/papers/research/finance.pdf)

Some more:

>- CNN asks: [Is democratic socialism the right path for America?](http://www.cnn.com/2015/10/27/opinions/opinion-roundup-socialism/)

@@

@@

> _"Congress shall make no law respecting an establishment of religion, or prohibiting the free exercise thereof; or abridging the freedom of speech, or of the press; or the right of the people peaceably to assemble, and to petition the Government for a redress of grievances"_

@@

> _"What is the general view that I will be examining? It is that, in the market for goods, government regulation is desirable whereas, in the market for ideas, government regulation is undersirable and should be strictly limited."_

@@

> _"The paradox is that government intervention which is so harmful in the one sphere becomes beneficial in the other."_

@@

> _"... it is usually those who press most strongly for an extension of government regulation in other markets who are most anxious for a vigorous enforcement of the First Amendment prohibitions on government regulation in the market for ideas"_

@@

> _"Because of the view that a free market in ideas is necessary to the maintenance of democratic institutions and , I believe, for other reasons also, intellectuals have shown a tendency to exalt the market for ideas and to depreciate the market for goods._ ___Such an attitude seems to me unjustified.___"

> _"I would put the point more bluntly. The market for ideas is the market in which the intellectual conducts his trade. The explanation of the paradox is self-interest and self-esteem. Self-esteem leads the intellectuals to magnify the importance of their own market. That others should be regulated seems natural, particularly as many of the intellectuals see themselves as doing the regulating. But self-interest combines with self-esteem to ensure that, while others are regulated, regulation should not apply to them."_

@@

> _"It is hard to believe that the general public is in a better position to evaluate competing views on economic and social policy than to choose between different kinds of food."_

> _"I do not believe that we will be able to form a judgement in which we can have any confidence unless we abandon the present ambivalence about the performance of government in the two markets and adopt a more consistent view."_

>- One might say a more "___coherent___" view

@@

```{r, echo=FALSE, out.width = "300px", out.height="400px"}
knitr::include_graphics("images/Nau.jpg")
```

@@

```{r, echo=FALSE, out.width = "300px", out.height="400px"}
knitr::include_graphics("images/deFinetti.jpg")
```

@@

> ___The argument that a rational person's degrees of belief must conform to the axioms of the probability calculus, since otherwise, by the Dutch book theorem, he would be vulnerable to a Dutch book.___

@@

> ___The proposition that anyone who (a) counts a bet on a proposition p as fair if the odds correspond to his degree of belief that p is true and who (b) is willing to make any combination of bets he would regard individually as fair will be vulnerable to a Dutch book provided his degrees of belief do not conform to the axioms of the probability calculus. Thus, anyone of whom (a) and (b) are true and whose degree of belief in a disjunction of two incompatible propositions is not equal to the sum of his degrees of belief in the two propositions taken individually would be vulnerable to a Dutch book.___

@@

```{r, echo=FALSE, out.width = "600px", out.height="490px"}
knitr::include_graphics("images/xkcd.png")
```

@@

> _"In this paper it will be argued that decision theory, game theory, and market theory can be placed on a more secure and parsimonious footing by appeal to the principle of_ ___no arbitrage___ _rather than the usual ordering axioms of preferences. The arbitrage princile already has a long history: in the literature on Bayesian statistics and decision theory, it was introduced by de Finetti (1937) as a basis for defining subjective probability... In the finance literature, the arbitrage principle explicitly underlies the capital structure theorem (Modigliani & Miller 1958), the valuation of cash flows (Ross 1978), and the pricing of assets and options (Black and Scholes 1973, Merton 1973, Ross 1976). No-arbitrage is central to classical welfare economics, where it goes by the name of 'Pareto optimality.'"_

@@

> _"It will be shown here that no-arbitrage is the core principle of rationality which remains after economic models are stripped of their excess baggage of intrinsically unverifiable or empirically disconfirmed assumptions about the perfectness of human cognition, particularly the assumption that belief and preference orderings are complete. The arbitrage principle still provides support for the most important concepts of equilibrium in games and markets, but with weaker information requirements. It is consistent with 'procedural' or 'cognitive' views of rationality (e.g., Simon 1976; Munier 1991) which emphasize that the information-processing capabilities of the individual are finite and tailored to the specific demands of her physical and cultural environment."_

@@

> _"Thus, Bayesians are made, not born."_

@@

>- Hayek points out that market processes are knowledge processes. 
>- It should seem very natural to economists to treat the market for ideas and the market for goods similarly. They are two different markets, but the same market process!
>- Thus, the process of scholarship and science are very much like the market process for goods.

@@

```{r, echo=FALSE, out.width = "600px", out.height="450px"}
knitr::include_graphics("images/thomas-sowell.jpg")
```

@@

> ___"While market economies are often thought of as money economies, they are still more so knowledge economies... Economic transactions are purchases and sales of knowledge."___

> ___"After all, the caveman had the same natural resources at their disposal as we have today... We are all in the business of buying and selling knowledge from one another, because we are each so profoundly ignorant of what it takes to complete the whole process of which we are a part."___

@@

## Notes

>- I am a bit embarrased to admit that until very recently I was not aware of Israel Kirzner's very important work on entrepreneurial arbitrage.
>- Dutch books in the election cycle? Like Thomas Sowell once said, "I feel like a mosquito at a nudist collony. Where do I begin?!"
>- Ben thinks that one of the problems is the lack of arbitragable positions. 
>- We all have the same vote, so there is no one person who can really move the market with superior knowledge. What we get is mass confusion! 
>- Once you start seeing the world this way you will see Dutch books everywhere. Especially on Facebook.

>>>>

---
title: "The Morality of Arbitrage"
author: "Tyler J. Brough, Ph.D."
date: "April 11, 2017"
output: slidy_presentation
---

# Introduction

##

@@

## About Me

- Financial Economist at Utah State University
- My research focuses on market microstructure, derivatives markets, financial econometrics, and computational methods.
- I'm on the [web](https://broughtj.github.io/)

<br>

<center>
<img src="images/buddy.jpg" width=400 height=350 />
</center>

@@

>- It is apparent to me that the standard ethical arguments for capitalism are not having any purchase in the minds of most people.
>- Capitalism is presented as some kind of Faustian bargain: we like it for its wealth production and efficiency, but at bottom it is really about greed. It is rapacious and immoral.
>- It is readily apparent that many of our colleagues in the business disciplines believe this notion
>- __I'm afraid that our students do too!__ They leave the business school with some good technical tools, but not with an appreciation of just how much good they will do in their careers. 
>- That is on us! That's a failure on our part to teach correct principles.

## Some Recent Anecdotal Evidence

@@

<center>
<img src="images/squid.jpg" width=400 height=350 />
</center>

## Sadly, Yet More

>- "Derivatives are financial weapons of mass destruction" - Warren Buffett
>- "Derivatives are toxic - stay away from them" - unnamed donor to the Huntsman School
>- "In a world where profit is consistently put before both people and the planet, climate economics has everything to do with ethics and morality" - Naomi Klein
>- CNN asks: [Is democratic socialism the right path for America?](http://www.cnn.com/2015/10/27/opinions/opinion-roundup-socialism/)

@@

- I can date my adult thinking about markets and the market process to my first reading of Hayek's 1945 AER article [*The Use of Knowledge in Society*](http://www.jstor.org/stable/1809376?seq=1#page_scan_tab_contents)

@@

> *"Congress shall make no law respecting an establishment of religion, or prohibiting the free exercise thereof; or abridging the freedom of speech, or of the press; or the right of the people peaceably to assemble, and to petition the Government for a redress of grievances"*

@@

> *"What is the general view that I will be examining? It is that, in the market for goods, government regulation is desirable whereas, in the market for ideas, government regulation is undersirable and should be strictly limited."*

@@

> *"The paradox is that government intervention which is so harmful in the one sphere becomes beneficial in the other."*

@@

> *"... it is usually those who press most strongly for an extension of government regulation in other markets who are most anxious for a vigorous enforcement of the First Amendment prohibitions on government regulation in the market for ideas"*

@@

> "*Because of the view that a free market in ideas is necessary to the maintenance of democratic institutions and , I believe, for other reasons also, intellectuals have shown a tendency to exalt the market for ideas and to depreciate the market for goods.* ***Such an attitude seems to me unjustified.***"

> *"I would put the point more bluntly. The market for ideas is the market in which the intellectual conducts his trade. The explanation of the paradox is self-interest and self-esteem. Self-esteem leads the intellectuals to magnify the importance of their own market. That others should be regulated seems natural, particularly as many of the intellectuals see themselves as doing the regulating. But self-interest combines with self-esteem to ensure that, while others are regulated, regulation should not apply to them."*

@@

> *"It is hard to believe that the general public is in a better position to evaluate competing views on economic and social policy than to choose between different kinds of food."*

> *"I do not believe that we will be able to form a judgement in which we can have any confidence unless we abandon the present ambivalence about the performance of government in the two markets and adopt a more consistent view."*

>- One might say a more "***coherent***" view

@@

<center>
<img src="images/Nau.jpg" width=400 height=350 />
</center>

@@

<center>
<img src="images/deFinetti.jpg" width=400 height=350 />
</center>

@@

> *The argument that a rational person's degrees of belief must conform to the axioms of the probability calculus, since otherwise, by the Dutch book theorem, he would be vulnerable to a Dutch book.*

@@

> *The proposition that anyone who (a) counts a bet on a proposition p as fair if the odds correspond to his degree of belief that p is true and who (b) is willing to make any combination of bets he would regard individually as fair will be vulnerable to a Dutch book provided his degrees of belief do not conform to the axioms of the probability calculus. Thus, anyone of whom (a) and (b) are true and whose degree of belief in a disjunction of two incompatible propositions is not equal to the sum of his degrees of belief in the two propositions taken individually would be vulnerable to a Dutch book.*

@@

<center>
<img src="images/xkcd.png" width=600 height=490 />
</center>

@@

> "*In this paper it will be argued that decision theory, game theory, and market theory can be placed on a more secure and parsimonious footing by appeal to the principle of* ***no arbitrage*** *rather than the usual ordering axioms of preferences. The arbitrage princile already has a long history: in the literature on Bayesian statistics and decision theory, it was introduced by de Finetti (1937) as a basis for defining subjective probability... In the finance literature, the arbitrage principle explicitly underlies the capital structure theorem (Modigliani & Miller 1958), the valuation of cash flows (Ross 1978), and the pricing of assets and options (Black and Scholes 1973, Merton 1973, Ross 1976). No-arbitrage is central to classical welfare economics, where it goes by the name of 'Pareto optimality.' *"

@@

> "*It will be shown here that no-arbitrage is the core principle of rationality which remains after economic models are stripped of their excess baggage of intrinsically unverifiable or empirically disconfirmed assumptions about the perfectness of human cognition, particularly the assumption that belief and preference orderings are complete. The arbitrage principle still provides support for the most important concepts of equilibrium in games and markets, but with weaker information requirements. It is consistent with 'procedural' or 'cognitive' views of rationality (e.g., Simon 1976; Munier 1991) which emphasize that the information-processing capabilities of the individual are finite and tailored to the specific demands of her physical and cultural environment.*"

@@

> "*Thus, Bayesians are made, not born.*"

@@

>- Hayek points out that market processes are knowledge processes, and that markets are information systems to use a modern parlance. 
>- It should seem very natural to economists to treat the market for ideas and the market for goods similarly. They are two different markets, but the same market process!
>- Thus, the process of scholarship and science are very much like the market process for goods.

@@

<center>
<img src="images/thomas-sowell.jpg" width=600 height=490 />
</center>

@@

> "*While market economies are often thought of as money economies, they are still more so knowledge economies... Economic transactions are purchases and sales of knowledge.*"

> "*After all, the caveman had the same natural resources at their disposal as we have today... We are all in the business of buying and selling knowledge from one another, because we are each so profoundly ignorant of what it takes to complete the whole process of which we are a part.*"

@@

## Conclusion

>- Thank you for your attention!
>- Your feedback is appreciated!
